---
title: "Reconstructing lineage tree with single cell RNA-Seq data"
author: "Qin Zhu, Kim Lab"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

## Dataset

`data/eset_ABprp.rds` contains the single cell expression data of the ABprp lineage, it is an ExpressionSet object.
`data/subtree_ABprp.rds` contains the corresponding lineage tree. 

```{r}
library(BiocGenerics)
library(Biobase)
library(Matrix)

cur_eset <- readRDS("data/eset_ABprp.rds")
cur_tree <- readRDS("data/subtree_ABprp.rds")
```

Now let's take a look at the ExpressionSet object:

```{r}
cur_eset@assayData$exprs[1:5, 1:5] # The exprs slot contains the raw count matrix in sparse matrix format
cur_eset@assayData$norm_exprs[1:5, 1:5] # The norm_exprs slot contains the log2 transformed normalized count matrix in sparse matrix format
head(pData(cur_eset)) # The pData() is used to get meta data for the cells, such as lineage, cell type, embryo time and batch
```

The tree table is shows the cell lineage relationship, from means parent cell, to means daughter cell of that parent cell, br_name means the birth time of the parent cell and d_time means the brith time of the daughter cell. "lineage" column is the most important column because it corresponds to the lineage annotation in the single cell data (`pData(cur_eset)$lineage`). Note for lineages like "ABpxpaaa", the "x" corresponds to "r" (right) or "l" (left) because of symmetry (ABprpaaa and ABplpaaa have same expression profile). In this ABprp subtree, you can treat it as "ABprpaaa".

```{r}
head(cur_tree)
```

With these two R object, the goal is to reconstruct the cell lineage tree (as presented in `cur_tree`) with just the single cell expression data (`cur_eset`). 

* Easy version of the problem: You can first compute average expression of each lineage using the `cur_eset@assayData$norm_exprs` and `pData(cur_eset)$lineage`, then try to reconstruct the lineage tree using the average expression. If you think average expression is not a good statistic to represent each lineage, you can come up with your own statistic, or directly build the tree with the single cell data.

* Complete version of the problem: Direct recontruct lineage tree from single cells.

## Visualization

Here's just some visualization to help you better understand the data. You do not need to understand any of the plotting function.

you can see in this plot, the lineage tree is shown on the top and the single cell data is shown at the bottom (presented as a UMAP plot). The links between the vertices of the lineage tree and the single cell clusters represents the mapping between the two.

```{r}
library(ggtree)
library(ggplot2)
library(dplyr)
library(tidytree)
source("scripts/plotFunctions.R")
source("scripts/tree_helper_function.R")
source("scripts/tree_plot_function.R")

top_tree <- make_lineage_ggtree(in_tree = cur_tree, root = "ABprp", time.cut = 250, color.annot = "lineage", branch.length='lifetime') + guides(color= F)
lineage_color = get_factor_color(unique(cur_eset$lineage), "Set1")
names(lineage_color) <- unique(cur_eset$lineage)
res <- make_tree_dimr(proj=pData(cur_eset)[, c("UMAP1", "UMAP2", "lineage")], left_tree = NULL, right_tree = NULL, top_tree = top_tree, shared.col = "lineage", colorBy = "lineage", tree.color = lineage_color, label.time.cut = 120,
                      plot.link = "top", shift.y.scale = 1/20, 
                      return_coords = F) 

res
```

## The Grand Challenge

This part is not required for your final project but you are encouraged to give it a try.

You can find the complete C. elegans embryogenesis dataset here: https://github.com/qinzhu/VisCello.celegans

Once you git clone `VisCello.celegans`, you can get the ExpressionSet containing all cells and the entire cell lineage tree from commands below:

```{r, eval=F}
# Assuming working directory being the VisCello.celegans
eset <- readRDS("inst/app/data/eset.rds") # Same formatted as eset above
lineage_table <- readRDS("inst/app/data/lineage_tree_tbl.rds") # Same as tree table above
```

Note there are some cells with ambiguous lineage mapping or no lineage mapping due to lack of information, in such case, you'll find the lineage of that cell as "unannotated". 

Can you reconstruct the entire C. elegans embryogenesis cell lineage tree with this data?









